---
title: "Calicoblast marker gene detectability under subsampling"
---

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(SeuratObject)
library(Matrix)
library(dplyr)
library(purrr)
library(ggplot2)
library(scales)
set.seed(42)
```

## Overview

We ask: at what cell count does the calicoblast marker gene signature become
undetectable? Coral datasets are subsampled to decreasing cell numbers and
UMI-downsampled to R. yuma's sequencing depth. The fraction of calicoblast
marker genes detectable (total UMI across calicoblast cells > 10) is recorded
across iterations. If marker genes remain detectable at R. yuma's cell count
(9,822), their absence in R. yuma reflects a genuine biological difference.

---

## 1  Parameters

```{r params}
out_dir        <- "/calicoblast_markers"
MARKER_DIR     <- "/calicoblast_markers"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

RYUMA_N_CELLS  <- 9822   # total cells in R. yuma
RYUMA_MEAN_UMI <- 885.3  # mean nCount_RNA per cell in R. yuma

N_ITER  <- 50   # iterations per cell-count point
SWEEP_N <- c(200, 500, 1000, 2000, 3000, 5000, 7000, 8000, 9000, RYUMA_N_CELLS, 11000,
             12000, 15000, 20000)

SPECIES_INFO <- list(
  Amil = list(
    path        = "/Amil.seurat_final.rds",
    ct_col      = "legacy_cell_type",
    cali_label  = "calicoblast",
    marker_file = file.path(MARKER_DIR, "Amil_calicoblast_markers.tsv")
  ),
  Spis = list(
    path        = "/Spis.seurat_final.rds",
    ct_col      = "legacy_cell_type",
    cali_label  = "calicoblast",
    marker_file = file.path(MARKER_DIR, "Spis_calicoblast_markers.tsv")
  ),
  Ocupat = list(
    path        = "/Ocupat.seurat_final.rds",
    ct_col      = "legacy_cell_type",
    cali_label  = "calicoblast",
    marker_file = file.path(MARKER_DIR, "Ocupat_calicoblast_markers.tsv")
  )
)

sp_order   <- c("Amil", "Spis", "Ocupat")
sp_colours <- c(Amil = "#E07B39", Spis = "#4A90D9", Ocupat = "#5BAD72")
```

---

## 2  Load data

```{r load-markers}
marker_list <- lapply(SPECIES_INFO, function(info) {
  df <- read.table(info$marker_file, sep = "\t", header = TRUE,
                   stringsAsFactors = FALSE)
  df$gene
})

cat("Marker gene counts:\n")
for (sp in names(marker_list)) cat(" ", sp, ":", length(marker_list[[sp]]), "\n")
```

```{r load-seurat, cache=TRUE, message=FALSE, warning=FALSE}
sobj_list <- lapply(SPECIES_INFO, function(info) readRDS(info$path))
names(sobj_list) <- names(SPECIES_INFO)

cat("Loaded species:\n")
for (sp in names(sobj_list)) {
  obj  <- sobj_list[[sp]]
  info <- SPECIES_INFO[[sp]]
  n    <- ncol(obj)
  nc   <- sum(obj@meta.data[[info$ct_col]] == info$cali_label, na.rm = TRUE)
  mu   <- mean(obj@meta.data[["nCount_RNA"]], na.rm = TRUE)
  nm   <- length(intersect(marker_list[[sp]], rownames(obj[["RNA"]])))
  cat(sprintf("  %-8s  %6d cells  %4d calicoblasts (%.1f%%)  mean UMI = %.0f  markers in matrix = %d\n",
              sp, n, nc, 100 * nc / n, mu, nm))
}
```

---

## 3  Helper functions

```{r helpers}
# Downsample each cell to at most `target` UMIs (sampling without replacement)
downsample_counts <- function(mat, target) {
  target      <- as.integer(round(target))
  mat         <- as(mat, "dgCMatrix")
  mat@x       <- as.numeric(round(mat@x))
  cell_totals <- Matrix::colSums(mat)
  cells_to_ds <- which(cell_totals > target)
  if (length(cells_to_ds) == 0) return(mat)

  ng <- nrow(mat)
  for (j in cells_to_ds) {
    col       <- mat[, j]
    nz        <- which(col > 0)
    if (length(nz) == 0) next
    molecules <- rep(nz, as.integer(col[nz]))
    if (length(molecules) <= target) next
    mat[, j]  <- tabulate(sample(molecules, target), nbins = ng)
  }
  mat
}

# One subsampling iteration: returns marker gene detection rate
# (fraction of marker genes expressed in >= 1 calicoblast cell)
one_iteration <- function(obj, info, markers, n_target, target_umi = RYUMA_MEAN_UMI) {
  # Cell subsampling
  idx      <- sample(ncol(obj), min(n_target, ncol(obj)))
  sub_meta <- obj@meta.data[idx, , drop = FALSE]
  cali_idx <- which(sub_meta[[info$ct_col]] == info$cali_label)

  if (length(cali_idx) == 0) return(NA_real_)

  cali_barcodes <- rownames(sub_meta)[cali_idx]
  counts_all    <- obj[["RNA"]]$counts
  valid_markers <- intersect(markers, rownames(counts_all))

  if (length(valid_markers) == 0) return(NA_real_)

  # UMI downsampling on calicoblast cells only
  cali_mat <- downsample_counts(
    counts_all[, cali_barcodes, drop = FALSE],
    target_umi
  )

  # Fraction of marker genes expressed in >= 1 calicoblast cell
  marker_mat       <- cali_mat[valid_markers, , drop = FALSE]
  n_detectable     <- sum(Matrix::rowSums(marker_mat) > 10) # Threshold: total UMI across all calicoblast cells > 10
  n_detectable / length(valid_markers)
}
```

---

## 4  Sweep analysis

```{r sweep}
sweep_cache <- file.path(out_dir, "sweep_v3_cache_UMI10.rds")

if (file.exists(sweep_cache)) {
  sweep_df <- readRDS(sweep_cache)
  cat("Loaded from cache.\n")
} else {
  results <- list()

  for (sp in names(sobj_list)) {
    cat("Processing", sp, "...\n")
    obj     <- sobj_list[[sp]]
    info    <- SPECIES_INFO[[sp]]
    markers <- marker_list[[sp]]
    max_n   <- ncol(obj)

    for (nc in SWEEP_N[SWEEP_N <= max_n]) {
      rates <- sapply(seq_len(N_ITER), function(i) {
        one_iteration(obj, info, markers, n_target = nc)
      })
      results[[paste0(sp, "__", nc)]] <- list(
        species           = sp,
        n_cells           = nc,
        mean_detection    = mean(rates, na.rm = TRUE),
        sd_detection      = sd(rates,   na.rm = TRUE),
        n_cali_zero_frac  = mean(is.na(rates))  # fraction of iters with 0 calicoblasts
      )
    }
  }

  sweep_df <- map_dfr(results, function(x) {
    tibble(
      species          = x$species,
      n_cells          = x$n_cells,
      mean_detection   = x$mean_detection * 100,
      sd_detection     = x$sd_detection   * 100,
      n_cali_zero_frac = x$n_cali_zero_frac
    )
  }) %>% mutate(species = factor(species, levels = sp_order))

  saveRDS(sweep_df, sweep_cache)
  cat("Done. Saved to", sweep_cache, "\n")
}

sweep_df
```

---

## 5  Figure

```{r fig-sweep, fig.width=6, fig.height=4}
p <- ggplot(sweep_df, aes(x = n_cells, y = mean_detection, colour = species,
                           fill = species)) +
  geom_errorbar(aes(ymin = pmax(mean_detection - sd_detection, 0),
                    ymax = pmin(mean_detection + sd_detection, 100)),
                width = 300,          
                linewidth = 0.5,
                alpha = 0.7) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 2) +
  geom_vline(xintercept = RYUMA_N_CELLS, linetype = "dashed",
             colour = "grey40", linewidth = 0.5) +
  annotate("text", x = RYUMA_N_CELLS * 1.04, y = 15,
           label = paste0("R. yuma\n(", comma(RYUMA_N_CELLS), " cells)"),
           hjust = 0, size = 2.8, colour = "grey30") +
  scale_x_continuous(labels = label_comma()) +
  scale_y_continuous(limits = c(0, 105), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_colour_manual(values = sp_colours) +
  scale_fill_manual(values = sp_colours) +
  labs(
    x      = "Cells subsampled",
    y      = "Calicoblast marker genes detected",
    colour = NULL,
    fill   = NULL,
    title  = "Marker gene detectability under subsampling",
    subtitle = paste0("UMI downsampled to R. yuma mean (", round(RYUMA_MEAN_UMI), "/cell)  |  ",
                      N_ITER, " iterations ± SD")
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "right") +
  scale_x_continuous(labels = label_comma(),
                   expand = c(0, 0)) +          
scale_y_continuous(limits = c(0, 105), breaks = seq(0, 100, 20),
                   labels = function(x) paste0(x, "%"),
                   expand = c(0, 0))

p
```

---

## 6  Save

```{r save}
pdf(file.path(out_dir, "calicoblast_marker_detectability_UMI10.pdf"), width = 6, height = 4)
print(p)
dev.off()

write.csv(sweep_df, file.path(out_dir, "calicoblast_marker_detectability_UMI10.csv"),
          row.names = FALSE)

cat("Saved to", out_dir, "\n")
```

---

## Session info

```{r session}
sessionInfo()
```
