---
title: "GO enrichment (topGO) — Calicoblast & Epidermis"
---

```{r libraries}
library(topGO)
library(GO.db)
library(AnnotationDbi)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
```

---

## Parameters

```{r params}
# GO ontology: "BP", "MF", or "CC"
ONTOLOGY <- "MF"

# topGO algorithm: "weight01" (topology-aware, recommended) or "classic"
ALGORITHM <- "weight01"

# p-value cutoff (topGO weight01 p-values are not BH-corrected by design;
# the weight01 algorithm itself penalises redundant parent terms)
P_CUT  <- 0.05
N_TOP  <- 10   # top terms per bar panel

# Directories
eggnog_coral_dir <- "/eggNOG"
marker_dir       <- "/calicoblast_markers"
out_dir          <- "/eggnog"

dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

species    <- c("Amil", "Spis", "Ocupat", "Gast")
cell_types <- c("calico", "epidermis")
```

---

## 1. Parse eggNOG annotation files

```{r parse-eggnog}
parse_eggnog <- function(path) {
  df <- read.table(
    path, sep = "\t", comment.char = "#",
    header = FALSE, stringsAsFactors = FALSE, quote = ""
  )
  df <- df[, c(1, 10)]
  colnames(df) <- c("query", "GOs")
  df[df$GOs != "-", ]
}

eggnog_raw <- list(
  Amil   = parse_eggnog(file.path(eggnog_coral_dir, "Amil_long.emapper.annotations")),
  Spis   = parse_eggnog(file.path(eggnog_coral_dir, "Spis_long.emapper.annotations")),
  Ocupat = parse_eggnog(file.path(eggnog_coral_dir, "Ocupat_long.emapper.annotations")),
  Gast   = parse_eggnog(file.path(eggnog_gast_dir,  "Gast.emapper.annotations"))
)

cat("Annotated genes per species:\n")
for (sp in species) cat(" ", sp, ":", nrow(eggnog_raw[[sp]]), "\n")
```

---

## 2. Build gene2GO lists (topGO format)

topGO requires a **named list**: `gene_id → character vector of GO IDs`.
Annotations are provided as direct (leaf) terms only — topGO propagates them
up the DAG internally.

```{r gene2go}
build_gene2go <- function(eggnog_df, sp) {
  # Strip isoform suffixes so IDs match marker gene format
  if (sp == "Amil")   eggnog_df$query <- sub("-RA$",  "", eggnog_df$query)

  go_list <- strsplit(eggnog_df$GOs, ",")
  names(go_list) <- eggnog_df$query
  go_list
}

gene2go_list <- lapply(setNames(species, species), function(sp) {
  build_gene2go(eggnog_raw[[sp]], sp)
})

cat("Universe size per species (genes with any GO annotation):\n")
for (sp in species) cat(" ", sp, ":", length(gene2go_list[[sp]]), "\n")
```

---

## 3. Load marker genes

```{r load-markers}
# Convert marker gene IDs to the same format as the eggnog query IDs
marker_to_eggnog <- function(genes, sp) {
  genes <- genes[!grepl("orphan", genes, ignore.case = TRUE)]

  if (sp == "Gast") {
    # g.TRINITY-GG-207-c43-g1-i1  →  TRINITY_GG_207_c43_g1_i1
    sub("^g\\.", "", gsub("-", "_", genes))
  } else if (sp == "Amil") {
    # Amil-Amillepora15760  →  Amil_Amillepora15760  (matches after -RA strip)
    gsub("-", "_", genes)
  } else if (sp == "Ocupat") {
    # Ocupat-v02-G022773  →  Ocupat_v02_G022773.t0
    paste0(gsub("-", "_", genes), ".t0")
  } else {
    # Spis-XP-022803870-1  →  Spis_XP_022803870_1
    gsub("-", "_", genes)
  }
}

epidermis_dir <- "/home/tosuke/work/Coral/Corallimorpharia/GO/R/output_epidermis_markers"

load_markers <- function(sp, ct) {
  if (ct == "calico") {
    path <- switch(sp,
      Amil   = file.path(marker_dir,    "Amil_calicoblast_markers.tsv"),
      Spis   = file.path(marker_dir,    "Spis_calicoblast_markers.tsv"),
      Ocupat = file.path(marker_dir,    "Ocupat_calicoblast_markers.tsv"),
      Gast   = file.path(marker_dir,    "Gast_calicoblast_markers.tsv")
    )
  } else {
    path <- switch(sp,
      Amil   = file.path(epidermis_dir, "Amil_epidermis_markers.tsv"),
      Spis   = file.path(epidermis_dir, "Spis_epidermis_markers.tsv"),
      Ocupat = file.path(epidermis_dir, "Ocupat_epidermis_markers.tsv"),
      Gast   = file.path(epidermis_dir, "Gast_epidermis_markers.tsv")
    )
  }
  df    <- read.table(path, sep = "\t", header = TRUE, stringsAsFactors = FALSE)
  genes <- df$gene
  marker_to_eggnog(genes, sp)
}

cat("Marker gene counts (after ID conversion):\n")
marker_list <- list()
for (sp in species) {
  for (ct in cell_types) {
    key <- paste(sp, ct, sep = "_")
    m   <- load_markers(sp, ct)
    marker_list[[key]] <- m
    cat(" ", key, ":", length(m), "\n")
  }
}
```

---

## 4. Run topGO enrichment

```{r topgo-fn}
run_topgo <- function(sp, ct) {
  key      <- paste(sp, ct, sep = "_")
  markers  <- marker_list[[key]]
  gene2go  <- gene2go_list[[sp]]
  all_genes <- names(gene2go)

  # Gene score vector: 1 = marker, 0 = not
  gene_scores <- factor(as.integer(all_genes %in% markers))
  names(gene_scores) <- all_genes

  n_sig <- sum(gene_scores == 1)
  if (n_sig < 3) {
    message(key, ": only ", n_sig, " markers in universe — skipping")
    return(NULL)
  }

  # Build topGOdata object
  go_data <- tryCatch(
    new("topGOdata",
        ontology  = ONTOLOGY,
        allGenes  = gene_scores,
        geneSel   = function(x) x == 1,
        annot     = annFUN.gene2GO,
        gene2GO   = gene2go,
        nodeSize  = 3          # minimum GO term size
    ),
    error = function(e) { message(key, " topGOdata error: ", e$message); NULL }
  )
  if (is.null(go_data)) return(NULL)

  # Run test
  test_result <- runTest(go_data, algorithm = ALGORITHM, statistic = "fisher")

  # Extract top results
  n_terms <- length(score(test_result))
  if (n_terms == 0) { message(key, ": no terms tested"); return(NULL) }

  tbl <- GenTable(
    go_data,
    pvalue   = test_result,
    orderBy  = "pvalue",
    topNodes = min(n_terms, 200),
    numChar  = 100
  )

  # GenTable returns p-values as character (e.g. "< 1e-30"); parse them
  tbl$pvalue <- as.numeric(sub("< ", "", tbl$pvalue))

  tbl <- tbl %>%
    filter(pvalue < P_CUT) %>%
    mutate(
      species   = sp,
      cell_type = ct,
      logP      = -log10(pvalue),
      Count     = Significant
    )

  if (nrow(tbl) == 0) { message(key, ": no terms pass p < ", P_CUT); return(NULL) }

  tbl
}

enrich_results <- lapply(species, function(sp) {
  lapply(cell_types, function(ct) run_topgo(sp, ct))
})
names(enrich_results) <- species
for (sp in species) names(enrich_results[[sp]]) <- cell_types

cat("\nSignificant terms per species/cell type:\n")
for (sp in species)
  for (ct in cell_types) {
    r <- enrich_results[[sp]][[ct]]
    cat(" ", sp, ct, ":", if (is.null(r)) "0" else nrow(r), "terms\n")
  }
```

---

## 5. Bar plots

```{r plot-fn}
clean_term <- function(term, width = 30) {
  lines <- strwrap(term, width = width)
  if (length(lines) > 2) paste0(paste(lines[1:2], collapse = "\n"), "...") else paste(lines, collapse = "\n")
}

truncate_term <- function(term, maxchar = 60) {
  ifelse(nchar(term) > maxchar, paste0(substr(term, 1, maxchar - 3), "..."), term)
}

make_bar <- function(sp, ct, n_top = N_TOP) {
  df <- enrich_results[[sp]][[ct]]

  if (is.null(df) || nrow(df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No enriched terms", size = 3) +
        theme_void() +
        labs(title = paste0(sp, " — ", ct))
    )
  }

  df <- df %>%
    slice_min(order_by = pvalue, n = n_top, with_ties = FALSE) %>%
    mutate(
      Term = sapply(Term, clean_term),
      Term = reorder(Term, logP)
    )

  x_max <- max(df$logP) * 1.25

  ggplot(df, aes(x = logP, y = Term)) +
    geom_col(fill = "white", color = "black", width = 0.65) +
    geom_text(aes(label = paste0("(", Count, ")")),
              hjust = -0.1, size = 2.8, colour = "black") +
    labs(
      title = paste0(sp, " — ", ct),
      x     = expression(-log[10](p-value)),
      y     = NULL
    ) +
    scale_x_continuous(expand = c(0, 0), limits = c(0, x_max)) +
    theme(
      panel.grid        = element_blank(),
      panel.background  = element_blank(),
      plot.background   = element_blank(),
      axis.line         = element_line(color = "black"),
      panel.border      = element_blank(),
      plot.title        = element_text(size = 10, face = "bold", colour = "black"),
      axis.text.y       = element_text(size = 8, colour = "black"),
      axis.text.x       = element_text(size = 8, colour = "black"),
      axis.title.x      = element_text(size = 8, colour = "black")
    )
}
```

### Bar plots: rows = species, columns = calicoblast | epidermis

```{r bar-grid, fig.width=8, fig.height=12}
plots <- lapply(species, function(sp) lapply(cell_types, function(ct) make_bar(sp, ct)))

bar_grid <- wrap_plots(
  plots[[1]][[1]], plots[[1]][[2]],
  plots[[2]][[1]], plots[[2]][[2]],
  plots[[3]][[1]], plots[[3]][[2]],
  plots[[4]][[1]], plots[[4]][[2]],
  ncol = 2
)
bar_grid
```

---

## 6. Bubble plots

```{r combine-results}
df_all <- do.call(rbind, lapply(species, function(sp) {
  do.call(rbind, lapply(cell_types, function(ct) enrich_results[[sp]][[ct]]))
}))
```

```{r bubble-fn}
make_bubble <- function(df, sp_levels, title_suffix = "") {
  if (is.null(df) || nrow(df) == 0)
    return(ggplot() + annotate("text", 0.5, 0.5, label = "No data") + theme_void())

  top_terms <- df %>%
    filter(species %in% sp_levels) %>%
    group_by(species, cell_type) %>%
    slice_min(pvalue, n = N_TOP, with_ties = FALSE) %>%
    ungroup() %>%
    pull(Term) %>%
    unique()

  plot_df <- df %>%
    filter(species %in% sp_levels, Term %in% top_terms) %>%
    mutate(
      Term      = truncate_term(Term),
      species   = factor(species,   levels = sp_levels),
      cell_type = factor(cell_type, levels = cell_types),
      logP      = -log10(pvalue)
    )

  term_mat <- plot_df %>%
    mutate(col_key = paste(species, cell_type, sep = "_")) %>%
    dplyr::select(Term, col_key, logP) %>%
    pivot_wider(names_from = col_key, values_from = logP,
                values_fill = 0, values_fn = max) %>%
    tibble::column_to_rownames("Term")

  if (nrow(term_mat) > 1) {
    hc         <- hclust(dist(term_mat))
    term_order <- rev(rownames(term_mat)[hc$order])
  } else {
    term_order <- rownames(term_mat)
  }

  plot_df$Term <- factor(plot_df$Term, levels = term_order)

  ggplot(plot_df, aes(x = species, y = Term, size = Count, fill = logP)) +
    geom_point(shape = 21, colour = "black") +
    scale_size_continuous(name = "Gene count", range = c(2, 8)) +
    scale_fill_gradient(
      name = expression(-log[10](p)),
      low = "white", high = "navy"
    ) +
    facet_grid(~ cell_type) +
    labs(x = NULL, y = NULL,
         title = paste0("GO enrichment (topGO ", ALGORITHM, ") — ", ONTOLOGY, title_suffix)) +
    theme_bw() +
    theme(
      axis.text.x      = element_text(size = 6, colour = "black", hjust = 0.5),
      axis.text.y      = element_text(size = 6, colour = "black"),
      axis.ticks       = element_line(colour = "black"),
      strip.text       = element_text(size = 6, face = "bold", colour = "black"),
      strip.background = element_rect(fill = "grey92", colour = "black"),
      legend.text      = element_text(size = 6, colour = "black"),
      legend.title     = element_text(size = 6, colour = "black"),
      panel.grid.major = element_line(colour = "grey90"),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(size = 6, face = "bold")
    )
}
```

### Bubble plot: all species

```{r bubble-all, fig.width=7, fig.height=13}
bubble_all <- make_bubble(df_all, c("Spis", "Ocupat", "Gast", "Amil"))
bubble_all
```

### Bubble plot: corals only (without Gast)

```{r bubble-no-gast, fig.width=7, fig.height=12}
bubble_no_gast <- make_bubble(df_all, c("Spis", "Ocupat", "Amil"), " (corals only)")
bubble_no_gast
```

---

## 7. Save outputs

```{r save}
pdf_bar <- file.path(out_dir, paste0("GO_topGO_", ONTOLOGY, "_bar.pdf"))
pdf(pdf_bar, width = 8, height = 12)
print(bar_grid)
dev.off()
cat("Saved:", pdf_bar, "\n")

pdf_bubble_all <- file.path(out_dir, paste0("GO_topGO_", ONTOLOGY, "_bubble_all.pdf"))
pdf(pdf_bubble_all, width = 7, height = 13)
print(bubble_all)
dev.off()
cat("Saved:", pdf_bubble_all, "\n")

pdf_bubble_ng <- file.path(out_dir, paste0("GO_topGO_", ONTOLOGY, "_bubble_corals.pdf"))
pdf(pdf_bubble_ng, width = 8, height = 10)
print(bubble_no_gast)
dev.off()
cat("Saved:", pdf_bubble_ng, "\n")

if (!is.null(df_all) && nrow(df_all) > 0) {
  csv_path <- file.path(out_dir, paste0("GO_topGO_", ONTOLOGY, "_results.csv"))
  write.csv(df_all, csv_path, row.names = FALSE)
  cat("Saved:", csv_path, "\n")
}
```

---

## Session info

```{r session}
sessionInfo()
```
