---
title: "TranscryptFormer: pan-calicoblast cosine similarity"
---

## Concept

Calicoblast cells from all species that carry this annotation (Amil, Spis, Ocup,
Gast) are pooled after per-species centering. The mean L2-normalised vector of
this pool is the **pan-calicoblast centroid**. We then compute its cosine
similarity against every cell-type centroid in every comparison species.

A high similarity to a specific cell type in a non-skeletal species suggests a
transcriptional homologue; a diffuse/low pattern suggests this cell identity is
novel to Scleractinia.

```{r libraries}
library(anndata)
library(ggplot2)
library(dplyr)
library(tidyr)
library(pheatmap)
library(RColorBrewer)
```

## Load embeddings

```{r load-data}
base_coral <- "/transcriptformer"
base_other <- "/transcriptformer_other"

adata_amil  <- read_h5ad(file.path(base_other, "Amil_inference/embeddings.h5ad"))
adata_spis  <- read_h5ad(file.path(base_other, "Spis_inference/embeddings.h5ad"))
adata_ocup  <- read_h5ad(file.path(base_other, "Ocupat_inference/embeddings.h5ad"))
adata_gast  <- read_h5ad(file.path(base_coral, "Gast_inference/embeddings.h5ad"))
adata_nve   <- read_h5ad(file.path(base_other, "Nve_inference/embeddings.h5ad"))
adata_xsp   <- read_h5ad(file.path(base_other, "Xsp_inference/embeddings.h5ad"))
adata_ryuma <- read_h5ad(file.path(base_coral, "Ryuma_inference/embeddings.h5ad"))

emb_amil  <- adata_amil$obsm[["embeddings"]]
emb_spis  <- adata_spis$obsm[["embeddings"]]
emb_ocup  <- adata_ocup$obsm[["embeddings"]]
emb_gast  <- adata_gast$obsm[["embeddings"]]
emb_nve   <- adata_nve$obsm[["embeddings"]]
emb_xsp   <- adata_xsp$obsm[["embeddings"]]
emb_ryuma <- adata_ryuma$obsm[["embeddings"]]

ct_amil  <- as.character(as.data.frame(adata_amil$obs)$cell_type)
ct_spis  <- as.character(as.data.frame(adata_spis$obs)$cell_type)
ct_ocup  <- as.character(as.data.frame(adata_ocup$obs)$cell_type)
ct_gast  <- as.character(as.data.frame(adata_gast$obs)$cell_type)
ct_nve   <- as.character(as.data.frame(adata_nve$obs)$cell_type)
ct_xsp   <- as.character(as.data.frame(adata_xsp$obs)$cell_type)
ct_ryuma <- as.character(as.data.frame(adata_ryuma$obs)$cell_type)
```

## Cluster merge maps

```{r merge-maps}
merge_ct <- function(labels, map) {
  result <- map[labels]
  unmapped <- is.na(result)
  result[unmapped] <- labels[unmapped]
  unname(result)
}

amil_map <- c(
  calicoblast                = "calicoblast",
  cnidocyte                  = "cnidocyte",
  digestive_filaments        = "digestive_filaments",
  epidermis                  = "epidermis",
  epidermis_like_1           = "epidermis",
  epidermis_like_2           = "epidermis",
  gastrodermis_1             = "gastrodermis",
  gastrodermis_2             = "gastrodermis",
  gastrodermis_like          = "gastrodermis",
  gastrodermis_mitotic       = "gastrodermis",
  gastrodermis_muscle        = "gastrodermis",
  gastrodermis_alga_hosting  = "alga_hosting_cell",
  germline_oocytes           = "germline",
  gland_1_Xbp                = "glands",
  gland_2                    = "glands",
  gland_3                    = "glands",
  gland_4                    = "glands",
  gland_5                    = "glands",
  gland_6                    = "glands",
  immune_like                = "immune_cell",
  immune_macrophage_like_1   = "immune_cell",
  immune_macrophage_like_2   = "immune_cell",
  neuron_Isl_1               = "neuron",
  neuron_Isl_2               = "neuron",
  neuron_Pou4                = "neuron",
  neuron_Pou4_Otp            = "neuron",
  neurosecretory_progenitors = "neuron"
)

spis_map <- c(
  calicoblast                = "calicoblast",
  cnidocyte                  = "cnidocyte",
  digestive_filaments        = "digestive_filaments",
  epidermis                  = "epidermis",
  gastrodermis               = "gastrodermis",
  gastrodermis_like_2        = "gastrodermis",
  gastrodermis_muscle        = "gastrodermis",
  gastrodermis_alga_hosting  = "alga_hosting_cell",
  germline_oocytes_1         = "germline",
  germline_oocytes_2         = "germline",
  gland_1_Xbp                = "glands",
  gland_2                    = "glands",
  gland_3                    = "glands",
  immune_macrophage          = "immune_cell",
  immune_macrophage_like_1   = "immune_cell",
  immune_macrophage_like_2   = "immune_cell",
  immune_nascent             = "immune_cell",
  neuron_Atoh8               = "neuron",
  neuron_Isl                 = "neuron",
  neuron_Isl_Nkx6            = "neuron",
  neuron_Pou4_1              = "neuron",
  neuron_Pou4_2              = "neuron",
  neuron_Pou4_Gsx            = "neuron",
  neuron_Pou4_Otp            = "neuron",
  neurosecretory_progenitors = "neuron"
)

ocup_map <- c(
  calicoblast                = "calicoblast",
  cnidocyte_1                = "cnidocyte",
  cnidocyte_2                = "cnidocyte",
  digestive_filaments        = "digestive_filaments",
  epidermis                  = "epidermis",
  epidermis_like_1           = "epidermis",
  epidermis_like_2           = "epidermis",
  gastrodermis               = "gastrodermis",
  gastrodermis_like_1        = "gastrodermis",
  gastrodermis_like_2        = "gastrodermis",
  gastrodermis_muscle_like   = "gastrodermis",
  gastrodermis_alga_hosting  = "alga_hosting_cell",
  germline_oocytes           = "germline",
  gland_1_Xbp                = "glands",
  gland_2                    = "glands",
  gland_3                    = "glands",
  gland_4                    = "glands",
  gland_5                    = "glands",
  gland_6                    = "glands",
  gland_7                    = "glands",
  gland_8                    = "glands",
  immune_like_1              = "immune_cell",
  immune_like_2              = "immune_cell",
  immune_macrophage_like     = "immune_cell",
  neuron_Isl                 = "neuron",
  neuron_Pou4_1              = "neuron",
  neuron_Pou4_2              = "neuron",
  neuron_Pou4_Gsx            = "neuron",
  neuron_Pou4_Otp_1          = "neuron",
  neuron_Pou4_Otp_2          = "neuron",
  neurosecretory_progenitors = "neuron"
)

gast_map <- c(
  "Algae_hosting_cells" = "alga_hosting_cell",
  "Calicoblast"         = "calicoblast",
  "Cnidocytes"          = "cnidocyte",
  "Epidermis_1"         = "epidermis",
  "Epidermis_2"         = "epidermis",
  "Gastrodermis"        = "gastrodermis",
  "Gland cells-like"    = "glands",
  "Immune cells"        = "immune_cell",
  "Neurons"             = "neuron",
  "unknown"             = "unknown"
)

nve_map <- c(
  cnidocyte_nematocytes                      = "cnidocyte",
  cnidocyte_spirocytes                       = "cnidocyte",
  digestive_filaments                        = "digestive_filaments",
  digestive_filaments_like                   = "digestive_filaments",
  epidermis                                  = "epidermis",
  epidermis_like_1                           = "epidermis",
  epidermis_like_2                           = "epidermis",
  gastrodermis                               = "gastrodermis",
  gastrodermis_muscle_parietal_circular_prog = "gastrodermis",
  gland_1                                    = "glands",
  gland_4                                    = "glands",
  gland_9                                    = "glands",
  gland_13                                   = "glands",
  gland_16                                   = "glands",
  gland_17                                   = "glands",
  gland_26                                   = "glands",
  gland_unknown_1                            = "glands",
  muscle                                     = "glands",
  neuron_10_vasopressin                      = "neuron",
  neuron_13                                  = "neuron",
  neuron_14_peptidergic                      = "neuron",
  neuron_16_peptidergic                      = "neuron",
  neuron_18                                  = "neuron",
  neuron_19                                  = "neuron",
  neuron_28_motor                            = "neuron",
  neuron_2_precursors                        = "neuron",
  neuron_30_peptidergic                      = "neuron",
  neuron_36_peptidergic                      = "neuron",
  neuron_4                                   = "neuron",
  neuron_7                                   = "neuron",
  neurosecretory_progenitors                 = "neuron",
  unknown                                    = "unknown"
)

xsp_map <- c(
  cnidocyte_1                = "cnidocyte",
  cnidocyte_2                = "cnidocyte",
  epidermis_1                = "epidermis",
  epidermis_2                = "epidermis",
  epidermis_3                = "epidermis",
  epidermis_like             = "epidermis",
  gastrodermis               = "gastrodermis",
  gastrodermis_gland_like    = "gastrodermis",
  gastrodermis_muscle        = "gastrodermis",
  gastrodermis_alga_hosting  = "alga_hosting_cell",
  gland_calicoblast_like     = "glands",
  gland_secretory_1          = "glands",
  gland_secretory_2          = "glands",
  gland_secretory_3          = "glands",
  gland_secretory_4          = "glands",
  gland_secretory_5          = "glands",
  neuron_Etv1                = "neuron",
  neuron_Hmx                 = "neuron",
  neuron_Pou4_1              = "neuron",
  neuron_Pou4_2              = "neuron",
  neuron_Pou4_3              = "neuron",
  neuron_Pou4_4              = "neuron",
  neurosecretory_progenitors = "neuron",
  precursors                 = "neuron"
)

ryuma_map <- c(
  "Algae_hosting_cells" = "alga_hosting_cell",
  "Cnidocytes"          = "cnidocyte",
  "Epidermis"           = "epidermis",
  "Gastrodermis_1"      = "gastrodermis",
  "Gastrodermis_2"      = "gastrodermis",
  "Immune cells"        = "immune_cell",
  "Neurons-like"        = "neuron",
  "Neurons_1"           = "neuron",
  "Neurons_2"           = "neuron",
  "Unknown"             = "unknown"
)
```

```{r apply-merges}
ct_amil_m  <- merge_ct(ct_amil,  amil_map)
ct_spis_m  <- merge_ct(ct_spis,  spis_map)
ct_ocup_m  <- merge_ct(ct_ocup,  ocup_map)
ct_gast_m  <- merge_ct(ct_gast,  gast_map)
ct_nve_m   <- merge_ct(ct_nve,   nve_map)
ct_xsp_m   <- merge_ct(ct_xsp,   xsp_map)
ct_ryuma_m <- merge_ct(ct_ryuma, ryuma_map)
```

## Build pan-calicoblast centroid

Each species' embeddings are **centred independently** before extracting
calicoblast cells, so that species-level global biases do not contaminate the
pan-calicoblast signal.

```{r pan-calicoblast}
centre <- function(emb) sweep(emb, 2, colMeans(emb))
l2_norm_vec <- function(v) v / sqrt(sum(v^2))

# Centre each species, then collect calicoblast cells
cali_sources <- list(
  Amil = list(emb = emb_amil, ct = ct_amil_m),
  Spis = list(emb = emb_spis, ct = ct_spis_m),
  Ocup = list(emb = emb_ocup, ct = ct_ocup_m),
  Gast = list(emb = emb_gast, ct = ct_gast_m)
)

cali_pool <- do.call(rbind, lapply(names(cali_sources), function(sp) {
  d   <- cali_sources[[sp]]
  emb_c <- centre(d$emb)
  n_cali <- sum(d$ct == "calicoblast")
  cat(sprintf("%-5s : %d calicoblast cells\n", sp, n_cali))
  emb_c[d$ct == "calicoblast", , drop = FALSE]
}))

cat(sprintf("\nTotal pooled calicoblast cells: %d\n", nrow(cali_pool)))

# Pan-calicoblast centroid: mean of all pooled cells, L2-normalised
pan_cali_centroid <- l2_norm_vec(colMeans(cali_pool))
```

## Helper: cell-type centroids for comparison species

```{r helpers}
mean_by_type_normed <- function(emb, labels) {
  emb_c <- centre(emb)
  types <- sort(unique(labels))
  mat <- do.call(rbind, lapply(types, function(ct)
    l2_norm_vec(colMeans(emb_c[labels == ct, , drop = FALSE]))))
  rownames(mat) <- types
  mat
}

# Species order: distant (no calicoblast) → close (has calicoblast)
comp_order <- c("Xsp", "Nve", "Ryuma", "Spis", "Ocup", "Gast", "Amil")
has_calicoblast <- c(Amil=TRUE, Spis=TRUE, Ocup=TRUE, Gast=TRUE,
                     Ryuma=FALSE, Xsp=FALSE, Nve=FALSE)

comp_centroids <- list(
  Amil  = mean_by_type_normed(emb_amil,  ct_amil_m),
  Spis  = mean_by_type_normed(emb_spis,  ct_spis_m),
  Ocup  = mean_by_type_normed(emb_ocup,  ct_ocup_m),
  Gast  = mean_by_type_normed(emb_gast,  ct_gast_m),
  Ryuma = mean_by_type_normed(emb_ryuma, ct_ryuma_m),
  Xsp   = mean_by_type_normed(emb_xsp,   ct_xsp_m),
  Nve   = mean_by_type_normed(emb_nve,   ct_nve_m)
)
```

## Cosine similarity: pan-calicoblast vs. each species

```{r compute-sim}
sim_df <- do.call(rbind, lapply(comp_order, function(sp) {
  centroids <- comp_centroids[[sp]]
  sims      <- as.numeric(centroids %*% pan_cali_centroid)
  data.frame(
    species    = sp,
    cell_type  = rownames(centroids),
    similarity = sims,
    is_cali    = rownames(centroids) == "calicoblast",
    has_cali   = has_calicoblast[sp],
    stringsAsFactors = FALSE
  )
}))

sim_df$species <- factor(sim_df$species, levels = comp_order)
```

## Visualisation

### Bar plots: pan-calicoblast similarity per species

```{r barplots, fig.width = 12, fig.height = 7}
ggplot(sim_df,
       aes(x = reorder(cell_type, similarity), y = similarity, fill = is_cali)) +
  geom_col() +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey40") +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
                    labels = c("FALSE" = "other", "TRUE" = "calicoblast"),
                    name = "Cell type") +
  facet_wrap(~ species, nrow = 2, scales = "free_x") +
  coord_flip() +
  labs(x = NULL, y = "Cosine similarity to pan-calicoblast centroid",
       title = "Pan-calicoblast cosine similarity vs. each cell type",
       subtitle = "Species order: Scleractinia → Corallimorpharia → Octocorallia → Actiniaria\nRed = calicoblast (if annotated in that species)") +
  theme_classic(base_size = 11) +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold"))
```

### Heatmap: all species × merged cell types

```{r heatmap, fig.width = 10, fig.height = 5}
merged_type_order <- c(
  "calicoblast", "epidermis", "gastrodermis", "alga_hosting_cell",
  "digestive_filaments", "glands", "immune_cell", "cnidocyte",
  "neuron", "germline", "unknown"
)

# Long-format data for ggplot heatmap
heat_df <- do.call(rbind, lapply(comp_order, function(sp) {
  d <- sim_df[sim_df$species == sp, ]
  data.frame(
    species   = sp,
    cell_type = merged_type_order,
    similarity = heat_mat[sp, merged_type_order],   # NA if absent
    stringsAsFactors = FALSE
  )
}))

# Build the full matrix for NA lookup (constructed above in the loop)
heat_mat <- matrix(
  NA_real_,
  nrow = length(comp_order),
  ncol = length(merged_type_order),
  dimnames = list(comp_order, merged_type_order)
)
for (sp in comp_order) {
  d <- sim_df[sim_df$species == sp, ]
  present <- intersect(d$cell_type, merged_type_order)
  heat_mat[sp, present] <- d$similarity[match(present, d$cell_type)]
}

heat_df <- do.call(rbind, lapply(comp_order, function(sp) {
  data.frame(
    species    = sp,
    cell_type  = merged_type_order,
    similarity = heat_mat[sp, merged_type_order],
    stringsAsFactors = FALSE
  )
}))
heat_df$species   <- factor(heat_df$species,   levels = comp_order)
heat_df$cell_type <- factor(heat_df$cell_type, levels = merged_type_order)
heat_df$is_na     <- is.na(heat_df$similarity)

# Species row label: mark those with calicoblast
sp_label_map <- setNames(
  ifelse(has_calicoblast[comp_order], paste0(comp_order, "*"), comp_order),
  comp_order
)
heat_df$sp_label <- factor(sp_label_map[as.character(heat_df$species)],
                            levels = sp_label_map[comp_order])

ggplot(heat_df, aes(x = cell_type, y = sp_label)) +
  geom_tile(aes(fill = similarity), colour = "white", linewidth = 0.5) +
#  geom_text(data = heat_df[!heat_df$is_na, ],
#            aes(label = sprintf("%.2f", similarity)),
#            size = 2.8, colour = "black") +
  scale_fill_gradientn(
    colours = colorRampPalette(c("navy", "white", "firebrick"))(100),
    limits  = c(-0.4, 0.4), oob = scales::squish,
    name    = "Cosine\nsimilarity", na.value = "grey"
  ) +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL,
       title = "Pan-calicoblast cosine similarity",
       subtitle = "* = species has calicoblast annotation") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0),
        axis.ticks   = element_blank(),
        panel.border = element_rect(colour = "grey", fill = NA))
```

### Top-matching cell type per species

```{r top-match, fig.width = 8, fig.height = 4}
top_df <- sim_df %>%
  group_by(species) %>%
  slice_max(similarity, n = 1) %>%
  ungroup() %>%
  mutate(label = sprintf("%s\n(%.2f)", cell_type, similarity))

ggplot(top_df, aes(x = species, y = similarity, fill = cell_type)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = label), vjust = -0.3, size = 3.2) +
  scale_fill_brewer(palette = "Set2", name = "Top cell type") +
  scale_y_continuous(limits = c(0, max(top_df$similarity) * 1.25),
                     name = "Cosine similarity") +
  labs(x = "Comparison species",
       title = "Highest-similarity cell type to pan-calicoblast centroid per species") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```
